---
title: "Life Expectancy & Health Spending: A Quarto Overview"
author: "Archana Kallepillil"
subtitle: "Exploring the Intersection of Life Expectancy, Infant Nutrition, and Government Health Spending"
date: "14-Apr-2024"
format: html  
---




>*Exploring the trajectory of global life expectancy from 1960 to 2022 reveals a compelling narrative. As life expectancy steadily rises, the analysis highlights the pivotal roles of infant nutrition and government spending*

### Echoes of Time: A Glimpse into 1960's Global Health Landscape

In 1960, amid significant social and economic shifts, the global health landscape revealed stark disparities. A world map plot depicting life expectancy across nations serves as our starting point, offering insight into the state of global health. In Africa and Asia, life expectancy figures were notably low, reflecting issues of underdevelopment, poverty, and limited healthcare access exacerbated by socio-political instabilities. In contrast, Europe and the Americas exhibited higher life expectancy rates, indicative of prosperity and advanced healthcare systems. Nonetheless, disparities persisted, particularly among marginalized communities, underscoring the global challenge of achieving equitable health outcomes amidst socio-economic divides.


```{r}
#| label: setup
#| include: false
#| echo: false
#| warning: false
#| message: false

#install packages required

install.packages("tidyverse")
install.packages("maps")
install.packages("gapminder")
install.packages("egg")
install.packages("plotly")

library(tidyverse)
library(readxl)
library(egg)
library(gapminder)
library(plotly)

# import datasets
unicef_metadata_raw <- read_excel("unicef_metadata_xl.xlsx")
unicef_ind_1_raw <- read_excel("unicef_indicator_1_xl.xlsx")
unicef_ind_2_raw <- read_excel("unicef_indicator_2_xl.xlsx")


# Data manipulations

unicef_metadata <- unicef_metadata_raw %>%
  select(country, alpha_3_code,
         time_period = year,
         population = `Population, total`,
         gdp_per_capita = `GDP per capita (constant 2015 US$)`,
         inflation = `Inflation, consumer prices (annual %)`, 
         life_expectancy = `Life expectancy at birth, total (years)`)

uni_ind1 <- unicef_ind_1_raw %>% 
  distinct(alpha_3_code, time_period, .keep_all = TRUE)

#filtering 2018, as the file has only 4 other years randomly
uni_ind2 <- unicef_ind_2_raw %>% 
  distinct(alpha_3_code, time_period, .keep_all = TRUE) %>%
  filter(time_period == 2018)


```

:::{.column-body}
```{r}
#| echo: false
#| warning: false
#| message: false 
#| fig.width: 12
#| fig.height: 6

# Map of life expectancy at different countries in year 1960
lifeExp_1960 <- unicef_metadata %>%
  filter(time_period == 1960)

map_world <- map_data("world")

map_lifeExp_1960 <- full_join(map_world, lifeExp_1960, by = c("region" = "country"))

p1 <- ggplot(map_lifeExp_1960) + 
  aes(long, lat, group = group, fill = life_expectancy) + 
  geom_polygon() +
  scale_fill_gradient(low = "darkblue", high = "lightblue", na.value = "grey") +
  theme_bw() + 
  labs(
    title = "Life Expectancy at birth across the world in 1960",
    caption = "Source: Unicef",
    x = "Latitude",
    y = "Longitude",
    fill = "Life Expectancy at Birth"
  )

ggarrange(p1, ncol = 1)
```

:::

### Investing in Tomorrow: Unraveling the Link Between Health Spending and Longevity

Delving deeper into the dynamics of health and public policy, let us now examine the link between government health spending in 2018 and life expectancy in 2019, utilizing a scatter plot with regression lines. This choice of data captures the time lag between healthcare investment and its impact on life expectancy. The visual representation illustrates the positive correlation between higher government health spending and greater life expectancy. Outliers, reflecting unique socio-economic or healthcare system characteristics, can also be identified. Overall, the analysis combines text insights with visual representation to highlight the complex relationship between government health spending and life expectancy, leveraging data from 2018 and 2019 to capture the dynamic nature of healthcare investments and their impact on population health outcomes.

:::{.column-body}
```{r}
#| echo: false
#| warning: false
#| message: false 
#| fig.width: 10
#| fig.height: 6

# Government expenditure in % of GDP, on healthcare in 2018 and its impact on life expectancy in 2019

cntry_continents <- gapminder %>%
  select(country, continent) %>%
  distinct


govExp_2018 <- uni_ind2 %>%
  filter(time_period == 2018) %>%
  mutate(NextYear = time_period + 1) %>%
  inner_join(unicef_metadata, by = c("alpha_3_code" = "alpha_3_code",
                                     "country" = "country", 
                                     "NextYear" = "time_period")) %>%
  inner_join(cntry_continents, by = c("country" = "country"))


p2 <- ggplot(data = govExp_2018) +
  aes(life_expectancy, obs_value,  color = continent) +
  geom_point() +
  geom_smooth() +
  theme_bw() + 
  # facet_wrap( ~ continent, nrow = 1) +
  labs(
    x = "Life Expectancy at Birth in Year 2019",
    y = "Govt expenditure on health (% of GDP) in Year 2018",
    title = "Correlation between Government Health Spending and Life Expectancy ",
    color = "Continents"
  )


ggarrange(p2, ncol = 1)
```

:::


### Nourishing Futures: Exploring Infant Nutrition Beyond Borders

Infant nutrition profoundly impacts lifelong health outcomes and national life expectancies. Across regions excluding Europe and the Americas, a bar graph highlights low consumption rates of essential egg/flesh foods among infants aged 6 to 23 months. This underscores the critical need for global interventions to address infant nutrition deficiencies, as inadequate nutrition during infancy can diminish population life expectancy. However, outliers exist, notably in populations adhering to vegan diets, where alternative sources of essential nutrients must be ensured. Thus, targeted efforts to enhance infant nutrition are imperative for improving both individual well-being and national life expectancies.

* **Infant Nutrition Vitality**: Essential for early childhood development and lifelong health outcomes.

* **Age Group Focus**: Concentrating on infants aged 6 to 23 months across regions excluding Europe and the Americas.

* **Bar Graph Insights**: Reveals low consumption rates of essential egg/flesh foods in many countries.



:::{.column-body}
```{r}
#| echo: false
#| warning: false
#| message: false 
#| fig.width: 12
#| fig.height: 6

# Food consumption

foodConsumption <- uni_ind1 %>%
  filter(sex == "Total") %>%
  inner_join(unicef_metadata, by = c("alpha_3_code" = "alpha_3_code",
                                     "country" = "country", 
                                     "time_period" = "time_period")) %>%
  inner_join(cntry_continents, by = c("country" = "country")) %>%
  group_by(country) %>%
  summarise(mean_food = mean(obs_value))

cols = rainbow(26, s=.6, v=.9)[sample(1:26,26)]

p3 <- ggplot(foodConsumption) +
  aes(country, mean_food, fill = country) +
  geom_col() +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  scale_fill_manual(values=cols) +
  labs(
    x = "Country",
    y = "Egg and/or flesh foods consumption in infants of 6-23 months",
    title = "Egg and/ or Flesh Food Consumption among infants/toddlers",
    color = "Countries"
  )


ggarrange(p3, ncol = 1)
```

:::
### Journey Through Time: Tracing the Evolution of Global Life Expectancy


In our exploration of global life expectancy trends from 1960 to 2022, we witness a remarkable story of steady improvement worldwide. Despite facing challenges such as disease outbreaks and socio-political upheavals, life expectancy has consistently risen, reflecting advancements in healthcare, technology, and living conditions. Particularly noteworthy is the significant progress achieved by countries with historically lower life expectancies, showcasing the effectiveness of targeted interventions and collaborative efforts. While celebrating these achievements, we acknowledge the ongoing need for continued dedication to improving global health outcomes, addressing disparities, and tackling emerging threats. By leveraging lessons from the past and embracing innovation, we can navigate towards a healthier, more equitable future for all.

:::{.column-body}
```{r}
#| echo: false
#| warning: false
#| message: false 


# Time series chart

#time series chart - life expectancy by country over the years

lifeExp_TimeSeries <- unicef_metadata %>%
  inner_join(cntry_continents, by = c("country" = "country")) %>%
  ggplot() +
  aes(time_period, life_expectancy, group = country, color = continent) +
  geom_line() +
  theme_bw() +
   labs(
    x = "Year",
    y = "Life Expectancy at Birth",
    title = "Evolution of Life Expectancy over the years (1960 to 2022)",
    color = "Continents"
  )

ggplotly(lifeExp_TimeSeries) %>%
  layout(width = 800, height = 450)


```

:::


As we draw the curtains on our exploration, we're reminded of the profound impact of government spending and infant nutrition on life expectancy. From the correlation between public health initiatives and increased longevity to the vital role of adequate nutrition in early life, the data offers compelling insights. Armed with this knowledge, let us chart a course towards a healthier, more equitable future for all.
